---
title: "Basic Techniques in R"
author: "Umair Sayeed"
fontsize: 12pt
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\part{}

\noindent 1. Suppose the following are the final high school GPAs of a group of
college students: 3.62, 3.84, 3.95, 4.00, 3.12, 2.86, 3.77, 2.53, 3.26, and 2.90.
Create a vector that stores these values in the order specified.  Choose a 
variable name that's descriptive but relatively short.
\vspace{0.2in}

```{r}
studentsGPAs <- c(3.62, 3.84, 3.95, 4.00, 3.12, 2.86, 3.77, 2.53, 3.26, 2.90)
studentsGPAs
```

\vspace{0.2in}
\noindent 2. Remove the $4^{\text{th}}$ value from the vector in problem 1 and 
store the new vector.  Then print the new vector.
\vspace{0.2in}

```{r}
newStudentsGPAs <- studentsGPAs[-4]
newStudentsGPAs

```

\vspace{0.2in}
\noindent 3. Add the GPA 2.14 to the end of the vector in problem 2 and store the 
new vector.  Then print it.
\vspace{0.2in}

```{r}
newStudentsGPAs <- c(newStudentsGPAs, 2.14)
newStudentsGPAs
```

\vspace{0.2in}
\noindent 4. Find the average, median, and standard deviation of the GPAs of this
group of students.
\vspace{0.2in}

```{r}
mean(newStudentsGPAs)
median(newStudentsGPAs)
sd(newStudentsGPAs)
```

\vspace{0.2in}
\noindent 5. Using an \texttt{R} function, find the length of the vector in 
problem 4.
\vspace{0.2in}

```{r}
length(newStudentsGPAs)
```

\vspace{0.2in}
\noindent 6. Suppose the following are the SAT quantitative scores for the same
group of students: 640, 680, 790, 730, 580, 560, 670, 450, 650, and 490, 
respectively.  Create and print a vector that stores these SAT scores in the 
order provided.
\vspace{0.2in}

```{r}
studentsSAT <- c(640, 680, 790, 730, 580, 560, 670, 450, 650, 490)
studentsSAT
```

\vspace{0.2in}
\noindent 7. Find the mean and median SAT quantitative scores for this group of
students.
\vspace{0.2in}

```{r}
mean(studentsSAT)
median(studentsSAT)
```

\vspace{0.2in}
\noindent 8. Create a vector with the following values (in this order): 1, 1, 1, 
0, 0, 0, 0, 0, 1, 1.  Suppose that 1 represents "male" and 0 represents "female."
\vspace{0.2in}

```{r}
boysAndGirls <- c(1, 1, 1, 0, 0, 0, 0, 0, 1, 1)
boysAndGirls

```

\vspace{0.2in}
\noindent 9. Find the average of the values in the gender vector you created in
problem 9.
\vspace{0.2in}

```{r}
mean(boysAndGirls)
paste("Percentage of males in the group: ", 100*mean(boysAndGirls), "%")
```

\vspace{0.2in}
\noindent 10. Change your gender vector to a factor.  Make sure the change is
permanent and print the result.
\vspace{0.2in}

```{r}
boysAndGirls_fac = as.factor(boysAndGirls)
boysAndGirls_fac
```

\vspace{0.2in}
\noindent 11. Create a data frame that contains three variables, which correspond
to the three vectors you've already created.  Then print the data frame.
\vspace{0.2in}

```{r}
my_df <- data.frame(
  boysAndGirls,
  studentsSAT,
  studentsGPAs
  
)

my_df
```

\vspace{0.2in}
\noindent 12. How many rows and columns does your data frame have?  Find these
using appropriate \texttt{R} function(s).
\vspace{0.2in}

```{r}
str(my_df)
nrow(my_df)
ncol(my_df)
```

\vspace{0.2in}
\noindent 13. Print the data associated with the $6^{\text{th}}$ student.
\vspace{0.2in}

```{r}
my_df[6,]
```


\part{}

\noindent 1. Create and store a matrix that has five rows and four columns and 
that contains the values 6, 3, 12, 7, 1, 2, 0, 25, 23, 19, 5, 2, 5, 7, 12, 11, 
10, 8, 9, and 5 (in that order).  Fill the matrix by column (the default).
\vspace{0.2in}

```{r}
myMat <- matrix(data = c(6, 3, 12, 7, 1, 2, 0, 25, 23, 19, 
                         5, 2, 5, 7, 12, 11, 10, 8, 9, 5),
                nrow = 5, ncol = 4) 
myMat

```

\vspace{0.2in}
\noindent 2. Print the data from the third column.
\vspace{0.2in}

```{r}
myMat[, 3]
```

\vspace{0.2in}
\noindent 3. Print the data from the fifth row.
\vspace{0.2in}

```{r}
myMat[5, ]
```

\vspace{0.2in}
\noindent 4. Remove the second and fourth rows from the matrix and store the 
resulting matrix using a different name.  Then print the new matrix.
\vspace{0.2in}

```{r}
myMatNew <- myMat[-c(2, 4),]
myMatNew
```


<!-- part 3 -->
\part{}

In this part you will use the \texttt{mtcars} dataset, which is included in base
\texttt{R}.

\vspace{0.2in}
\noindent 1. Use the \texttt{str()} function to examine the structure of the 
data set.
\vspace{0.2in}

```{r}
str(mtcars)
```

\vspace{0.2in}
\noindent 2. Print the first six rows of the dataset.
\vspace{0.2in}

```{r}
mtcars[1:6,]
```

\vspace{0.2in}
\noindent 3. Print the mpg's (\texttt{mpg}) of all of the cars in the data set 
using subsetting.
\vspace{0.2in}

```{r}
mtcars$mpg
```

\vspace{0.2in}
\noindent 4. Find the average and median mpg of the cars in the data set.
\vspace{0.2in}

```{r}
mean(mtcars$mpg)
median(mtcars$mpg)
```

\vspace{0.2in}
\noindent 5. Filter the cars that got less than 15 mpg.
\vspace{0.2in}

```{r}
mtcars[mtcars$mpg < 15, ]  
```

\vspace{0.2in}
\noindent 6. Filter the cars with an mpg of at least 20 and a horsepower under 
100.
\vspace{0.2in}

```{r}
mtcars[mtcars$mpg >= 20 & mtcars$hp < 100, ]  
```

\vspace{0.2in}
\noindent 7. Remove the second, third, sixth, tenth, and fifteenth rows from the 
\texttt{mtcars} data frame and store the result.  Then print the new data frame.
\vspace{0.2in}

```{r}
mtcarsNew <- mtcars[-c(2, 3, 6, 10, 15),]
mtcarsNew
```

\vspace{0.2in}
\noindent 8. Print the miles per gallon (\texttt{mpg}) and horsepower 
(\texttt{hp}) variables by subsetting the \texttt{mtcars} data frame.
\vspace{0.2in}

```{r}
mpg_hp_df <- data.frame(mtcars$mpg, mtcars$hp)
mpg_hp_df

```


<!-- part 4 -->
\part{}

In this part you will work with data sets from the website for the Lock, 
Lock, Lock, Lock, and Lock textbook \textit{Statistics: Unlocking the Power of
Data}. Go to the following URL to access these data sets: 
\url{http://www.lock5stat.com/datapage.html}.
\vspace{0.1in}


\noindent 1. Download the CSV file for the \textit{Cars2015} data set and read it
into \texttt{R}. Use \texttt{str()} to examine the structure of the data set. 
Then make three plots: one for city mpg (\texttt{CityMPG}), one for car type 
(\texttt{Type}), and one for the city mpg based on the car type. Make sure you're 
using an appropriate plot for each (i.e., a histogram or boxplot for a 
quantitative variables and a bar graph for a categorical variable).
\vspace{0.2in}

```{r}
Cars2015 <- read.csv(file = "C:/Users/sayee/Downloads/Cars2015.csv",
                     header = TRUE)  

str(Cars2015)  

hist(Cars2015$CityMPG)  

barplot(table(Cars2015$Type))  

plot(x = as.factor(Cars2015$Type), y = Cars2015$CityMPG)  

```

\vspace{0.2in}
\noindent 2. Download the CSV file for the \textit{HollywoodMovies} data set. As
you look at the data set, do you notice any missing values?  If so, keep track of
what symbol(s) they are represented by.  Read your data into \texttt{R}, making
sure you use the \texttt{na.strings} argument, where \texttt{na.strings = 
[something]}. I'll leave it to you to figure out what should go on the right-hand 
side of the equal sign.

Now use the \texttt{summary()} function to find the five-number summary (as well
as the mean) for the Rotten Tomatoes score variable (\texttt{RottenTomatoes}). 
Note that this function also prints the number of NAs, if any exist.
\vspace{0.2in}

```{r}
HollywoodMovies <- read.csv(file = "C:/Users/sayee/Downloads/HollywoodMovies.csv",
                            header = TRUE, na.strings = "")  

summary(HollywoodMovies$RottenTomatoes)  

```

\vspace{0.2in}
\noindent 3. Search for the text (.txt) file for the \textit{ACS} data set.  Copy
and paste the entire data set into a notepad/wordpad/textedit file (whatever your
computer has for basic text files).  Then read the data into \texttt{R}, making 
sure any missing values are represents by NAs.  Next, make a scatterplot of the 
number of hours worked per week (\texttt{HoursWk}) by age (\texttt{Age}) for the 
individuals in the data set; i.e., age should be on the x-axis and the number of 
hours worked per week on the y-axis.
\vspace{0.2in}

```{r}
ACS <- read.table(file = "C:/Users/sayee/Downloads/ACS.txt",header = TRUE, sep = "\t")
plot(x = ACS$Age, y = ACS$HoursWk)
```



