---
title: "Linear Regression"
author: "Umair Sayeed"    
fontsize: 12pt
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\part{}

Use the \texttt{NJ\_Homes} dataset.  It contains data related to NJ housing prices from 2018.  Suppose 
you want to model the selling price of a house based on the size of the house 
(measured in square feet).
\vspace{0.2in}

\noindent 1. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}


\vspace{0.2in}
```{r}
"Reponse variable: Selling Price of House ($) = Quantitative
Predictor variable: Size of House (square feet) = Quantitative"
```

\vspace{0.2in}
\noindent 2. Create a scatterplot with the correct variable on each axis and with 
the regression line added to the plot.
\vspace{0.2in}

```{r}

homes <- read.csv(file = "C:/Users/sayee/Downloads/NJ_Homes.csv",
                  header = TRUE)
lm_info <- lm(PRICE ~ SQFT, data = homes)
coeffs <- lm_info$coefficients
plot(homes$SQFT, homes$PRICE, pch = 19, cex.lab = 1.2, cex.axis = 1, 
     cex = 1, xlab = "Size of House (square feet)", 
     ylab = "Selling Price of House ($)")
abline(a = coeffs[1], b = coeffs[2], lwd = 1)
```

\vspace{0.2in}
\noindent 3. Is linear regression an appropriate tool to use to model selling 
price based on size?  Why or why not?
\vspace{0.2in}


\vspace{0.2in}
```{r}
"Linear regression requires a linear relationship 
 between variables, where the dependent/response
 variable is quantitative, hence its use is appropriate
 to model a home's selling price based on size here."

```

\vspace{0.2in}
\noindent 4. What is the regression equation?
\vspace{0.2in}


\vspace{0.2in}
```{r}
lm_info
"predicted price = slope*(house size in sq. ft.) + y-intercept"
"Y = 261.4x - 26339"
```

\vspace{0.2in}
\noindent 5. Interpret the regression coefficients.  If it isn't appropriate to 
interpret either of them, explain why.
\vspace{0.2in}
\vspace{0.2in}
```{r}
lm_info
"Slope: For every unit increase in the value
 of the predictor variable, house size (sq. ft.),
 the price of the house will increase by $261.4321
 on average.
 Y-Int.: It is not appropriate to interpret the
 y-intercept because X = Size of House (square feet) = 0
 is not a possible condition for houses."

```

\vspace{0.2in}
\noindent 6. Predict the selling price of a 2,750 square foot house in NJ.  First 
do so by showing all of your work.  Then verify your answer using the 
\texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}


\vspace{0.2in}
```{r}
Y <- 261.4*2750 - 26338.9514
Y
predict(object = lm_info, newdata = data.frame(SQFT = 2750))
"A house that covers 2750 sq. ft. is predicted to cost $692599.4."
```

\vspace{0.2in}
\noindent 7. A 2,750 square foot house in NJ just went on the market.  Use a 95\% 
prediction interval to predict the selling price of the house.  Interpret your 
result and comment on its usefulness for predicting.  Note that the lowerbound
should be negative here.
\vspace{0.2in}


\vspace{0.2in}
```{r}
predict(object = lm_info, newdata = data.frame(SQFT = 2750), interval = "prediction")
"We predict with 95% confidence that the price of a house
 that covers 2750 sq. ft. is between $-308778.6 and $1693977.
 The prediction interval in this case is not very useful for
 predicting because the interval is so large at 95% confidence."
```

\vspace{0.2in}
\noindent 8. Suppose there's a 2,300 square foot house in the dataset that sold 
for $\$800,000$.  Find the residual for that house.  Was the selling price over-
or under-predicted?  By how much?
\vspace{0.2in}


\vspace{0.2in}
```{r}
predict(object = lm_info, newdata = data.frame(SQFT = 2300))
"actual Y - predicted Y = $800000 - $574955 = Residual = $225045. 
 The selling-price was underpredicted by $225045."
```

\vspace{0.2in}
\noindent 9. Find and interpret the residual standard error.
\vspace{0.2in}


\vspace{0.2in}
```{r}
summary(lm_info)
"The price of houses sold given their area in sq. ft.
 deviates from the regression line by $507900 on average."
```


\part{}

Use the \texttt{iqsize} dataset.  It contains data related to IQ and physical characteristics of a
group of people.  Suppose you want to model IQ (\texttt{PIQ}) based on the size 
of a person's brain (\text{Brain}).
\vspace{0.2in}

\noindent 1. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Reponse variable: PIQ = Quantitative
Predictor variable: Brain (units of volume) = Quantitative"

```

\vspace{0.2in}
\noindent 2. Create a scatterplot with the correct variable on each axis and with 
the regression line added to the plot.
\vspace{0.2in}

```{r}
  brains <- read.table(file = "C:/Users/sayee/Downloads/iqsize.txt",
                    header = TRUE)
  lm_info <- lm(PIQ ~ Brain, data = brains)
  coeffs <- lm_info$coefficients
  plot(brains$Brain, brains$PIQ, pch = 19, cex.lab = 1.2, cex.axis = 1, 
       cex = 1, xlab = "Size of Brain (units of volume)", 
       ylab = "IQ")
  abline(a = coeffs[1], b = coeffs[2], lwd = 1)

```

\vspace{0.2in}
\noindent 3. Is linear regression an appropriate tool to use to model IQ based on 
brain size?  Why or why not?
\vspace{0.2in}

```{r}
"Since the response variable is quantitative and there doesn't
 appear to be a clear curve present, linear regression is justified
 as an appropriate tool to model IQ based on brain size."

```

\vspace{0.2in}
\noindent 4. What is the regression equation?
\vspace{0.2in}


\vspace{0.2in}
```{r}
lm_info
"predicted iq = slope*(size of brain) + y-intercept"
"Y = 1.177x + 4.652"

```

\vspace{0.2in}
\noindent 5. Interpret the regression coefficients.  If it isn't appropriate to 
interpret either of them, explain why.
\vspace{0.2in}

```{r}
lm_info
"Slope: For every unit increase in brain size (units of volume), 
 PIQ increases by 1.177 IQ points.
 Y-Int.: It would not be appropriate to interpret this coefficient 
 because X = Size of Brain (vol. in units) = 0 
 is not a possible condition."
```

\vspace{0.2in}
\noindent 6. Predict the IQ of a person with a brain size of 95 units.  First do 
so by showing all of your work.  Then verify your answer using the 
\texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}


\vspace{0.2in}
```{r}
Y <- 1.177*95 + 4.652
Y
predict(object = lm_info, newdata = data.frame(Brain = 95))
"A brain that is 95 cubic units is predicted to have an IQ of 116.43."
```

\vspace{0.2in}
\noindent 7. Use a 95\% prediction interval to predict the IQ of a person with a
brain size of 95 units.  Interpret your result and comment on its usefulness for
predicting.
\vspace{0.2in}


\vspace{0.2in}
```{r}
predict(object = lm_info, newdata = data.frame(Brain = 95), interval = "prediction")
"We predict with 95% confidence that the IQ of a brain that is 95 cubic units
 in volume is between 72.64535 and 160.2147 IQ points. 
 A vast majority of people fall between an IQ score of 70 and 130,
 so the prediction interval in this case is not very useful for
 predicting because the interval is so large."

```

\vspace{0.2in}
\noindent 8. Suppose there's a person with an IQ of 100 and a brain size of 105
units.  Find the residual for that person.  Was their IQ over- or 
under-predicted?  By how much?
\vspace{0.2in}


\vspace{0.2in}
```{r}
predict(object = lm_info, newdata = data.frame(Brain = 105))
"actual Y - predicted Y = 100 - 128.1961 = Residual = -29.1961. 
 The selling-price was overpredicted by 29.1961 IQ points."

```

\vspace{0.2in}
\noindent 9. Find and interpret the residual standard error.
\vspace{0.2in}


\vspace{0.2in}
```{r}
summary(lm_info)
"The IQ of a brain given its volume deviates from
 the regression line by 21.21 IQ points on average."

```



