---
title: "RegressionAndResiduals"
author: "Umair Sayeed"   
fontsize: 12pt
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 6, 
  fig.align = "center"
)
```

\part{}

Use the \texttt{Homes\_Sales} dataset.  It contains data related to NJ housing prices from 2018.  Suppose 
you want to model the selling price of the house (\texttt{PRICE}) based on the 
number of bedrooms (\texttt{BEDS}) and the size of the house (\texttt{SQFT}, 
measured in square feet).
\vspace{0.2in}


\noindent 1. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Predictor variable: BEDS (Quantitative), SQFT (Quantitative)"
"Response variable: PRICE (Quantitative)"

```

\vspace{0.2in}
\noindent 2. What is the regression equation?  Write it two ways -- first using
$x$'s and $y$'s and then using the variable names.
\vspace{0.2in}



\vspace{0.2in}
```{r}
"PRICE = -201538.1 + 140.7*SQFT + 122764.8*BEDS"
"Y-hat = -201538.1 + 140.7*X1 + 122764.8*X2"
```

\vspace{0.2in}
\noindent 3. Create a residual plot.  Based on the residual plot, is the linear 
model you fit to the data reasonable to use to model the selling price of the 
house based on the number of bedrooms and the size of the house?  Why or why 
not?  Note: you should notice a couple of outliers, but they don't seem to have 
a strong impact on the model here.  Regardless of your answer, continue answering 
the rest of the questions.
\vspace{0.2in}



\vspace{0.2in}
```{r}
homes <- read.csv(file = "C:/Users/sayee/Downloads/Home_Sales.csv", 
                  header = TRUE)
homes.lm <- lm(PRICE ~ SQFT + BEDS, data = homes)
summary(homes.lm)
homes.res = resid(homes.lm)

coeffs <- homes.lm$coefficients
coeffs
plot(homes.lm$fitted.values, homes.res, 
    ylab = "Residuals", xlab = "Price ($)")
abline(0,0)
```

\vspace{0.2in}
\noindent 4. Interpret the regression coefficients.  If it isn't appropriate to 
interpret any of them, explain why.
\vspace{0.2in}

```{r}
"SQFT Slope: For every additional square foot of area, 
 holding the # of bedrooms constant, the home appreciates
 in value by $140.68 on average."
"BEDS Slope: For every additional bedroom in a home, 
 holding the area (sqft.) constant, the home appreciates
 in value by $122764.83 on average"
"Intercept: It is inappropriate to interpret the intercept
 because there are no homes in the dataset that meet the
 conditions: SQFT = 0 and BEDS = 0."

```

\vspace{0.2in}
\noindent 5. Predict the selling price of a 2,500 square foot house with three 
bedrooms.  First do so by showing all of your work.  Then verify your answer 
using the \texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
PRICE = -201538.1 + 140.6808*2500 + 122764.8297*3
PRICE
predict(object = homes.lm, newdata = data.frame(SQFT = 2500, BEDS = 3))
"The selling price of a 2500 sqft. house with 3
 bedrooms is predicted to be $518458.4."
```

\vspace{0.2in}
\noindent 6. Predict the selling price of an 1,800 square foot house with two 
bedrooms.  First do so by showing all of your work.  Then verify your answer 
using the \texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
PRICE = -201538.1 + 140.6808*1800 + 122764.8297*2
PRICE
predict(object = homes.lm, newdata = data.frame(SQFT = 1800, BEDS = 2))
"The selling price of an 1800 sqft home with 2 bedrooms
 is predicted to be $297217"

```

\vspace{0.2in}
For the rest of Part I, suppose you want to model the selling price of the house 
(\texttt{PRICE}) based on the same variables as before (the number of bedrooms,
\texttt{BEDS}, and the size of the house \texttt{SQFT}), as well as the number of 
bathrooms (\texttt{BATHS}).
\vspace{0.2in}

\noindent 7. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Predictor variable: BEDS (Quantitative), SQFT (Quantitative),
                     BATHS (Quantitative)"
"Response variable: PRICE (Quantitative)"
```


\vspace{0.2in}
\noindent 8. What is the regression equation?  Write it two ways -- first using
$x$'s and $y$'s and then using the variable names.
\vspace{0.2in}


\vspace{0.2in}

```{r}
"PRICE = -234630.46 + 48.64*SQFT + 59085.18*BEDS + 170510.49*BATHS"
"Y-hat = -234630.46 + 48.64*X1 + 59085.18*X2 + 170510.49*X3"

```

\vspace{0.2in}
\noindent 9. Create a residual plot.  Based on the residual plot, is the linear 
model you fit to the data reasonable to use to model the selling price of the 
house (based on the number of bedrooms, the number of bathrooms, the size of the 
lot, and the size of the house?  Why or why not?  Note: you should notice an 
outlier, but it doesn't seem to have a strong impact on the line here. Regardless 
of your answer, continue answering the rest of the questions.
\vspace{0.2in}



\vspace{0.2in}
```{r}
homes <- read.csv(file = "C:/Users/sayee/Downloads/Home_Sales.csv", 
                  header = TRUE)
homes.lm <- lm(PRICE ~ SQFT + BEDS + BATHS, data = homes)
summary(homes.lm)
homes.res = resid(homes.lm)

coeffs <- homes.lm$coefficients
coeffs
plot(homes.lm$fitted.values, homes.res, 
     ylab = "Residuals", xlab = "Price ($)")
abline(0,0)

"The linear model fit to the data is reasonable to use to model the selling
 price of a home based on the number of bedrooms, number of bathrooms,
 and the size of the house (sqft) because the residuals are roughly balanced."
```

\noindent 10. Predict the selling price of a 2,500 square foot house with three 
bedrooms and two bathrooms.  First do so by showing all of your work.  Then
verify your answer using the \texttt{predict()} function in \texttt{R}. 
Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
PRICE = -234630.46 + 48.64*2500 + 59085.18*3 + 170510.49*2
PRICE
homes.lm <- lm(PRICE ~ SQFT + BEDS + BATHS, data = homes)
predict(object = homes.lm, newdata = data.frame(SQFT = 2500,
                                                BEDS = 3, BATHS = 2))
"The selling price of a 2500 sqft home with 3 bedrooms
 and 2 bathrooms is predicted to be $405256.7."
```

\vspace{0.2in}
\noindent 11. Predict the selling price of an 1,800 square foot house that has 
two bedrooms and two bathrooms.  First do so by showing all of your work.  Then 
verify your answer using the \texttt{predict()} function in \texttt{R}. 
Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
PRICE = -234630.46 + 48.64*1800 + 59085.18*2 + 170510.49*2
PRICE
homes.lm <- lm(PRICE ~ SQFT + BEDS + BATHS, data = homes)
predict(object = homes.lm, newdata = data.frame(SQFT = 1800,
                                                BEDS = 2, BATHS = 2))
"The selling price of an 1800 sqft home with 2 bedrooms
 and 2 bathrooms is predicted to be $312120.5."

```

\vspace{0.2in}
\noindent 12. Suppose there's a 2,350 square foot house with four bedrooms and 
three bathrooms that sold for $\$575,000$.  Find the residual for that house. 
Was the house's selling price over- or under-predicted?  By how much?
\vspace{0.2in}



\vspace{0.2in}
```{r}
predict(object = homes.lm, newdata = data.frame(SQFT = 2350,
                                                BEDS = 4, BATHS = 3))
"The selling price was overpredicted by $52555.7."
```




\part{}

Use the \texttt{birthsmokers} dataset. It contains data related to births (of humans).  Suppose you 
want to model the birth weight of the baby (\texttt{Wgt\_Lbs}, in pounds) based 
on the length of the gestation period (\texttt{Gest}, in weeks).
\vspace{0.2in}


\noindent 1. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Predictor variable: Gest (Quantitative)"
"Response variable: Wgt_Lbs (Quantitative)"

```

\vspace{0.2in}
\noindent 2. Create a scatterplot with the correct variable on each axis and with 
the regression line added to the plot.  Based on the scatterplot, is the linear 
model you fit to the data reasonable to use to model the birth weight of the baby
based on the length of the gestation period?  Why or why not?
\vspace{0.2in}



\vspace{0.2in}
```{r}
birthsmokers <- read.table("C:/Users/sayee/Downloads/birthsmokers.txt", 
                          header = TRUE)
birthsmokers.lm <- lm(Wgt_Lbs ~ Gest, data = birthsmokers)
birthsmokers.lm

coeffs <- birthsmokers.lm$coefficients
coeffs

plot(birthsmokers$Gest, birthsmokers$Wgt_Lbs, pch = 19, 
     cex.lab = 1, cex.axis = 1, cex = 1,
     xlab = "Gestation Period (weeks)", 
     ylab = "Birth Weight (lbs)")

abline(a = coeffs[1], b = coeffs[2], lwd = 2)

"The linear model that was fit to the data is reasonable to use to
 model birth weight based on the length of the gestation period (weeks)
 because the datapoints follow a linear trend, not a curve trend."

```

\vspace{0.2in}
\noindent 3. Create a residual plot.  Based on the residual plot, is the linear 
model you fit to the data reasonable to use to model the birth weight of the baby
based on the length of the gestation period?  Why or why not?
\vspace{0.2in}

 

\vspace{0.2in}
```{r}
birthsmokers.res = resid(birthsmokers.lm)
plot(birthsmokers.lm$fitted.values, birthsmokers.res, 
     ylab = "Residuals", xlab = "Birth Weight (lbs)")
abline(0,0)

"The linear model fit to the data is reasonable to use to model
 birth weight (lbs) based on the length of the gestation period
 (weeks) because the residuals are balanced."

```

\vspace{0.2in}
\noindent 4. What is the regression equation?  Write it two ways -- first using
$x$'s and $y$'s and then using the variable names.
\vspace{0.2in}



\vspace{0.2in}
```{r}
summary(birthsmokers.lm)
"Wgt_Lbs = -4.49082 + 0.28840*Gest"
"Y-hat = -4.49082 + 0.28840*X"
```

\vspace{0.2in}
\noindent 5. Interpret the regression coefficients.  If it isn't appropriate to 
interpret either of them, explain why.
\vspace{0.2in}

```{r}

"Slope: For every additional week of gestation, the
 birth weight increases by 0.28840 lbs on average."
"Intercept: It is not appropriate to interpret the 
 intercept because there are no items in the dataset
 that satisfy the condition: Gest = 0."

```

\vspace{0.2in}
\noindent 6. Predict the birth weight of a baby for whom the length of the 
gestation period was 40 weeks.  First do so by showing all of your work.  Then 
verify your answer using the \texttt{predict()} function in \texttt{R}. 
Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
weight = -4.49082 + 0.28840*40
weight
predict(object = birthsmokers.lm, newdata = data.frame(Gest = 40))
"Given a gestation period of 40 weeks, the baby's
 birth weight is predicted to be 7.045216 lbs."
```

\vspace{0.2in}
\noindent 7. Use a 95% prediction interval to predict the birth weight of a baby 
for whom the length of the gestation period was 40 weeks.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
predict(object = birthsmokers.lm, newdata = data.frame(Gest = 40),
        interval = "prediction")
"We predict with 95% confidence that birth weight
 is between 6.276207 lbs and 7.814224 lbs given
 a gestation period of 40 weeks."

```

\vspace{0.2in}
For the rest of Part II, suppose you want to model the birth weight of the baby
(\texttt{Wgt\_Lbs}, in pounds) based on the length of the gestation period 
(\texttt{Gest}, in weeks) as well as the mother's smoking status (\texttt{Smoke}, 
where 1 indicates the mother was a smoker and 0 indicates the mother was not a 
smoker).
\vspace{0.2in}

\noindent 8. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Predictor variable: Smoke (Qualitative), Gest (Quantitative)"
"Response varable: Wgt_Lbs (Quantitative)"


```

\vspace{0.2in}
\noindent 9. What is the regression equation?  Write it two ways -- first using
$x$'s and $y$'s and then using the variable names.
\vspace{0.2in}



\vspace{0.2in}
```{r}
birthsmokers.lm <- lm(Wgt_Lbs ~ Gest + Smoke, data = birthsmokers)
birthsmokers.lm
"Wgt_Lbs = -5.2681 +0.3155*Gest - 0.5391*Smoke"
"Y-hat = -5.2681 +0.3155*X1 - 0.5391*X2"
```

\vspace{0.2in}
\noindent 10. Interpret the regression coefficients.  If it isn't appropriate to 
interpret any of them, explain why.
\vspace{0.2in}

```{r}
"Gest Slope: Holding the Smoke variable constant, every 
 additional week of gestation causes increases birth weight
 by 0.3155 lbs on average."
"Smoke Slope: Holding the gestation period constant, the birth
 weight of a baby whose mother was a smoker is 0.5391 lbs lighter
 than a baby whose mother was a smoker, on average."
"Intercept: It is not appropriate to interpret the intercept because
 there are no elements in the dataset that satisfy the condition Gest = 0."

```

\vspace{0.2in}
\noindent 11. Predict the birth weight of a baby for whom the length of the 
gestation period was 40 weeks and whose mother was a smoker.  First do so by 
showing all of your work.  Then verify your answer using the \texttt{predict()} 
function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
Wgt = -5.2681 +0.3155*40 - 0.5391
Wgt
predict(object = birthsmokers.lm, newdata = data.frame(Gest = 40,
                                                       Smoke = 1))
"Given a gestation period of 40 weeks, the
 birth weight of a baby whose mother was a 
 smoker is predicted to be 6.812042 lbs."
```

\vspace{0.2in}
\noindent 12. Predict the birth weight of a baby for whom the length of the 
gestation period was 40 weeks and whose mother was \underline{not} a smoker. 
First do so by showing all of your work.  Then verify your answer using the 
\texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
wgt = -5.2681 +0.3155*40 - 0.5391*0
wgt
predict(object = birthsmokers.lm, newdata = data.frame(Gest = 40,
                                                       Smoke = 0))
"Given a gestation period of 40 weeks, the
 birth weight of a baby whose mother was not 
 a smoker is predicted to be 7.351169 lbs."
```

\vspace{0.2in}
\noindent 13. Suppose a baby was born weighing 7.2 pounds to a mother who was a
smoker and after a gestation period lasting 37 weeks.  Find the residual for that 
baby.  Was the baby's birth weight over- or under-predicted?  By how much?
\vspace{0.2in}



\vspace{0.2in}
```{r}
predict(object = birthsmokers.lm, newdata = data.frame(Gest = 37,
                                                       Smoke = 1))
"The baby's birth was underpredicted by 1.019987 lbs"
```



\part{}

Use the \texttt{NELS} dataset.  It's a sample from a larger nationally representative longitudinal 
dataset that measures achievement outcomes in four core subject areas: math, 
science, reading, and history.  It also contains personal, familial, social, 
institutional, and cultural factors that might relate to the achievement 
outcomes.  Suppose you want to model student math achievement scores in 12th 
grade (\texttt{achmat12}) based on socioeconomic status (\texttt{ses}), the 
region of the country in which the student attends school (\texttt{region}), and 
gender (\texttt{gender}).  Note that the \texttt{region} coding is as follows: 1 
represents the northeast, 2 the north central region, 3 the south, and 4 the 
west.  The \texttt{gender} coding is as follows: 0 represents male and 1 female.
\vspace{0.2in}


\noindent 1. What are the response and predictor variables?  What type of 
variable is each -- quantitative or qualitative?
\vspace{0.2in}

```{r}
"Predictor variables: gender (Qualitative), region (Qualitative),
 ses (Quantitative)"
"Response variable: achmat12 (Quantitative)"
```


\vspace{0.2in}
\noindent 2. What is the regression equation?  Write it two ways -- first using
$x$'s and $y$'s and then using the variable names.  Make sure all qualitative 
variables are factors in \texttt{R} (all just to be safe)!
\vspace{0.2in}



\vspace{0.2in}
```{r}
"achmat12 = 52.3298955 + 0.3491868*ses - 0.1492133*region - 2.7346526*gender"
"Y-hat = 52.3298955 + 0.3491868*X1 - 0.1492133*X2 - 2.7346526*X3"
```

\vspace{0.2in}
\noindent 3. Create a residual plot.  Based on the residual plot, is the linear 
model you fit to the data reasonable to use to model student math achievement 
scores in $12^{\text{th}}$ grade based on socioeconomic status, the region of the 
country in which the student attends school, and gender?  Regardless of your
answer, continue answering the questions.
\vspace{0.2in}



\vspace{0.2in}

```{r}
nels <- read.csv(file = "C:/Users/sayee/Downloads/NELS.csv", 
                  header = TRUE)
nels$region <- as.factor(nels$region)
nels$gender <- as.factor(nels$gender)
is.factor(nels$region)
is.factor(nels$gender)
levels(nels$region)
levels(nels$gender)

nels.lm <- lm(achmat12 ~ ses + region + gender, data = nels)
summary(nels.lm)
nels.res = resid(nels.lm)

coeffs <- nels.lm$coefficients
coeffs
plot(nels.lm$fitted.values, nels.res, 
     ylab = "Residuals",
     xlab = "Math Achievement Scores in 12th Grade")
abline(0,0)

"The linear model fit to the data is reasonable to use to model
 student math achievement scores in 12th grade based on socioeconomic
 status, region of the country in which the sudent attends school, and
 gender because the residuals are roughly balanced."
```

\vspace{0.2in}
\noindent 4. Predict the math achievement score for a female student in 
$12^{\text{th}}$ grade with a socioeconomic status of 28 and who lives in the 
northeast.  First do so by showing all of your work.  Then verify your answer 
using the \texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
score = 52.3298955 + 0.3491868*28 - 0.1492133 - 2.7346526
score
predict(object = nels.lm, newdata = data.frame(ses = 28, region = "1",
                                               gender = "1"))
"The math achievement score for a female student in 12th grade
 at a school in the northeast with a socioeconomic status of 28
 is predicted to be 59.67395"
```

\vspace{0.2in}
\noindent 5. Predict the math achievement score for a male student in 
$12^{\text{th}}$ grade with a socioeconomic status of 28 and who attends school 
in the west.  First do so by showing all of your work.  Then verify your answer 
using the \texttt{predict()} function in \texttt{R}.  Interpret your result.
\vspace{0.2in}



\vspace{0.2in}
```{r}
score = 52.3298955 + 0.3491868*28 - 0.1492133*4 - 2.7346526*0
score
predict(object = nels.lm, newdata = data.frame(ses = 28, region = "4",
                                               gender = "0"))
"The math achievement score for a male student in 12th grade
 at a school in the west with a socioeconomic status of 28
 is predicted to be 62.24763."
```

\vspace{0.2in}
\noindent 6. Suppose a female student who attends school in the north central 
region has a socioeconomic status of 15 and a $12^{\text{th}}$-grade math 
achievement score of 65.  Find the residual for that student.  Was her 
$12^{\text{th}}$-grade math achievement score over- or under-predicted?  By how 
much?
\vspace{0.2in}



\vspace{0.2in}
```{r}
predict(object = nels.lm, newdata = data.frame(ses = 15, region = "2",
                                               gender = "1"))
"The student's math achievement score was underpredicted by 10.46538."
```



